Co-moving stars in \textsl{Gaia DR1}

### 2016-11-20 paper/ms.tex
Improve likelihood ratio figure and apply style
- New version of likelihood ratio figure
- apply consistent style and labels to all figures

\begin{abstract}
% Context
The primary sample of the \gaia\ \textsl{First Data Release} is the brand-new
\textsl{Tycho-Gaia Astrometric Solution} (\tgas). The precision and size of this sample
presents an opportunity to find new binary stars and moving groups, especially
rare binaries or sparsely populated groups.
% Aims
Here, we look for high-confidence co-moving pairs of stars using this data.
% Methods
We use a justified marginalized likelihood ratio test to separate
pairs of stars with similar three-space velocities from
those consistent with being drawn independently from the field
population.  Although we perform some visualizations using a
(bias-corrected) inverse parallax as a distance, the likelihood test
works in the observable space and uses the \gaia\ noise model
responsibly.
% Results
We find pairs of co-moving stars out to very wide
separations up to 10~pc, the limit of our search.
Some of these pairs form larger networks through mutual co-moving neighbors,
many of which correspond to the known open clusters, and stars in OB associations.
However, there still is a surprisingly large number (\todo{SMOH: give number})
of very wide ($>3$~pc) separation pairs of stars that are co-moving.
We present a catalog of high-confidence co-moving pairs of stars in \tgas\
with auxiliary information,
and discuss its potential usage
for making dynamical inferences, and testing stellar models and chemical
abundance measurements.

@@

  \item  However, it is not only (wide) binaries, which are gravitationally
  bound, that are expected to appear as co-moving.
  OB associations, moving groups, and open clusters are also of consideration.
  Since the basic unit of our choice is a pair of stars, we can expect that
  these astrophysical objects may be detected as networks of pairs.
  As the pair separation increases, the nature of co-moving pairs
  will change from binaries to those related to these larger objects,
  which generally subtend a larger angle in sky.
  Since the proper motions of two stars with the same 3D velocity
  are essentially projections of this velocity onto sphere at
  two different viewing angles,
  the larger the difference in viewing angles is, the larger the difference in tangential
  velocities will be.
  Due to this projection effect, a population of genuine co-moving pairs
  will extend to larger \absdvtan\ at larger separation.
  This indeed can be seen in Figure~\ref{fig:dv-sep} as an over-density
  in the lower right corner that gets thinner as \absdvtan\ increases.

  \item Finally, there is a population of “random” pairs of field stars
  which are not actually co-moving, but still have $\absdvtan < 10$~\kms\
  by chance.
  As \absdvtan\ increases, this population will dominate.
  However, it is clear from this Figure that there is an overlap between
  genuine co-moving pairs and “random” pairs.
\end{itemize}

@@

We compute the fully marginalized likelihood (FML) that a given pair of stars
has the same 3-space velocity with a small velocity difference drawn from a
zero-mean Gaussian with variance $s^2$ (hypothesis 1, $\mathcal{L}_1$), and
the FML of the stars having independent 3-space velocities (hypothesis 2,
$\mathcal{L}_2$).
We use the FML ratio $\mathcal{L}_1/\mathcal{L}_2$ as a scalar for selecting
candidate co-moving stars, as described in $\sectionname~\ref{sub:selection}$ in
more detail.
To compute these FMLs, the likelihood functions for each star in a pair, $L_i,
L_j$, are marginalized over true 3-space velocity and distance for each star in
the pair $(i,j)$.
\begin{align}
  \mathcal{L}_1 &=
    \int \, \dd d_i \, \dd d_j \, \dd^3 \vec{v} \,
    L_i(\vec{v}, d_i, s^2) \,
    L_j(\vec{v}, d_j, s^2) \,
    p(\vec{v}) \, p(d_i \given \varpi_i) \, p(d_j \given \varpi_j) \\
  \mathcal{L}_2 &=
    \int \, \dd d_i \, \dd d_j \, \dd^3 \vec{v}_i \, \dd^3 \vec{v}_j \,
    L_i(\vec{v}_i, d_i, 0) \,
    L_j(\vec{v}_j, d_j, 0) \,
    p(\vec{v}_i) \, p(\vec{v}_j) \, p(d_i \given \varpi_i) \, p(d_j \given \varpi_j). \label{eq:hyp2}
\end{align}
The marginalization integral for hypothesis 2 can be split into the product of
two simpler integrals $\mathcal{L}_2 = Q_i \, Q_j$ where
\begin{equation}
  Q = \int \, \dd d \, \dd^3 \vec{v} \, L(\vec{v}, d, 0) \, p(\vec{v}) \, p(d\given\varpi)
\end{equation}
If the velocity prior $p(\vec{v})$ is also Gaussian, the integrals over velocity
in both cases can be performed analytically:
We use a mixture of three isotropic, zero-mean Gaussian distributions
\begin{equation}
  p(\vec{v}) = \sum_{m=1}^3 \, w_m \mathcal{N}(0, \sigma_{v,m}^2)
\end{equation}
with velocity dispersions $(\sigma_{v,1}, \sigma_{v,2}, \sigma_{v,3}) = (15, 30, 50)
~\kms$ and weights $(w_1,w_2,w_3) = (0.3, 0.55, 0.15)$
meant to encompass young thin disk stars to halo stars.
We derive the relevant expressions in Appendix~\ref{sec:appendix}.
After marginalizing over velocity, the likelihood integrands only depend on
distance; we numerically compute the integrals over the true distances to each
star in a pair, $d_1,d_2$, using Monte Carlo integration with $K$ samples from
the distance posterior pdfs \todo{Comment on the choice of distance prior}:
\begin{equation}
  \int \, \dd d \, \tilde{L}(d) \, p(d\given\varpi) \approx
    \frac{1}{K} \, \sum_k^K \, \tilde{L}(d_k)
\end{equation}
where $\tilde{L}(d)$ is the velocity-marginalized likelihood function.
Through experimentation, we have found that $K=128$ samples are sufficient for
estimating the above integrals for stars with a wide range in parallax
signal-to-noise.
%TODO: SMOH: still unsure about this

@@

\begin{figure*}[htbp]
  \begin{center}
    \includegraphics[width=\textwidth]{figures/likelihoodratios.pdf}
  \end{center}
  \caption{%
    \todo{
    Left: Density histogram of likelihood ratios for candidate pairs and random pairs.
    Along with the distribution for the entire candidate pairs (black thick line),
    we show the distribution for subsets in slices of $\Delta v_t$.
    Right: $\Delta v_t$ vs separation in slices of $\ln \mathcal{L}_1 /\mathcal{L}_2$.}
    \label{fig:likelihoodratios}}
    \vspace{1em}
\end{figure*}

@@

Once we have identified candidate co-moving star pairs from the initial pair
sample, these pairs form an undirected graph where stars are nodes, and edges
between the nodes exist for co-moving pairs of stars.
A star may have multiple co-moving neighbors, and two stars may either
be directly or indirectly connected by a set of edges.
We divide the network into connected components\footnote{A connected component of an undirected graph $G$
is a subgraph of $G$ in which any two nodes are connected to each other by a path.
We use the term ``network'' loosely to refer to a graph or its connected components.},
and show the distribution of their sizes in Figure~\ref{fig:hist_ccsize}.
The most common are size 2 networks, which mean mutually exclusive co-moving
pairs. However, it is clear that there are many aggregates of co-moving stars
discovered by looking for co-moving pairs.
These aggregates are likely moving groups, OB associations, or star clusters.
There are 4555 connected components among 10,606 unique stars.
The maximum size of connected components is 151.
We show this largest network along with four randomly selected
examples of networks with sizes greater than 4 in Figure~\ref{fig:graphviz_examples}.
The largest network corresponds to the Pleiades open cluster.
One particular star in this network, TYC 1799-272-1, has 94 directly connected
co-moving neighbors, and is located at the center of the network.

@@

\begin{figure*}[htbp]
  \begin{center}
    \includegraphics[width=\textwidth]{figures/gjg_mepairs.png}
  \end{center}
  \caption{
    Color-magnitude diagrams of stars in networks of size 2 (i.e., mutually exclusive)
    in two separation bins.
    We connect each pair by a line on the left for pairs with separations smaller than
    1~pc.
    We show the distribution of all stars common in \tgas\ and \tmass\
    with parallax $S/N>16$ in blue for reference same as in Figure~\ref{fig:cmd_large}.
    \label{fig:cmd_me}}
\end{figure*}

>>>>

\begin{abstract}
% Context
The primary sample of the \gaia\ \textsl{First Data Release} is the brand-new
\textsl{Tycho-Gaia Astrometric Solution} (\tgas). The precision and size of this sample
presents an opportunity to find new binary stars and moving groups, especially
rare binaries or sparsely populated groups.
% Aims
Here, we look for high-confidence co-moving pairs of stars using this data.
% Methods
We use a justified marginalized likelihood ratio test to separate
pairs of stars with similar three-space velocities from
those consistent with being drawn independently from the field
population.  Although we perform some visualizations using a
(bias-corrected) inverse parallax as a distance, the likelihood test
works in the observable space and uses the \gaia\ noise model
responsibly.
% Results
We find pairs of co-moving stars out to very wide
separations up to 10~pc, the limit of our search.
Some of these pairs form larger networks through mutual co-moving neighbors,
many of which correspond to the known open clusters, and stars in OB associations.
However, there still is a surprisingly large number
of very wide ($>3$~pc) separation pairs of stars that are co-moving.
We present a catalog of high-confidence co-moving pairs of stars in \tgas\
with auxiliary information,
and discuss its potential usage
for making dynamical inferences, and testing stellar models and chemical
abundance measurements.

@@

  \item  A sample of co-moving stars also include stars that are part of OB associations,
    moving groups, and open clusters.
    These astrophysical objects may be detected as a network of co-moving pairs.
    As the pair separation increases, the nature of co-moving pairs
    will change from binaries to those related to these larger objects,
    which generally subtend a larger angle in sky.
    Since the proper motions of two stars with the same 3D velocity
    are essentially projections of this velocity onto sphere at
    two different viewing angles,
    the larger the difference in viewing angles is, the larger the difference in tangential
    velocities will be.
    Due to this projection effect, a population of genuine co-moving pairs
    will extend to larger \absdvtan\ at larger separation.
    This indeed can be seen in Figure~\ref{fig:dv-sep} as an over-density
    in the lower right corner that gets thinner as \absdvtan\ increases.

  \item Finally, there is a population of “random” pairs of field stars
    that are not co-moving, but still have $\absdvtan < 10$~\kms\
    by chance.
    As \absdvtan\ increases, this population will dominate.
    Figure~\ref{fig:dv-sep} shows that there is an overlap between
    genuine co-moving pairs and “random” pairs.
\end{itemize}

@@

We compute the fully marginalized likelihood (FML) that a given pair of stars
has the same 3-space velocity with a small velocity difference drawn from a
zero-mean Gaussian with variance $s^2$ (hypothesis 1, $\mathcal{L}_1$), and
the FML of the stars having independent 3-space velocities (hypothesis 2,
$\mathcal{L}_2$).
We use the FML ratio $\mathcal{L}_1/\mathcal{L}_2$ as a scalar for selecting
candidate co-moving stars, as described in $\sectionname~\ref{sub:selection}$ in
more detail.
To compute these FMLs, the likelihood functions for each star in a pair, $L_i,
L_j$, are marginalized over true 3-space velocity and distance for each star in
the pair $(i,j)$.
\begin{align}
  \mathcal{L}_1 &=
    \int \, \dd d_i \, \dd d_j \, \dd^3 \vec{v} \,
    L_i(\vec{v}, d_i, s^2) \,
    L_j(\vec{v}, d_j, s^2) \,
    p(\vec{v}) \, p(d_i \given \varpi_i) \, p(d_j \given \varpi_j) \\
  \mathcal{L}_2 &=
    \int \, \dd d_i \, \dd d_j \, \dd^3 \vec{v}_i \, \dd^3 \vec{v}_j \,
    L_i(\vec{v}_i, d_i, 0) \,
    L_j(\vec{v}_j, d_j, 0) \,
    p(\vec{v}_i) \, p(\vec{v}_j) \, p(d_i \given \varpi_i) \, p(d_j \given \varpi_j). \label{eq:hyp2}
\end{align}
The marginalization integral for hypothesis 2 can be split into the product of
two simpler integrals $\mathcal{L}_2 = Q_i \, Q_j$ where
\begin{equation}
  Q = \int \, \dd d \, \dd^3 \vec{v} \, L(\vec{v}, d, 0) \, p(\vec{v}) \, p(d\given\varpi)
\end{equation}
If the velocity prior $p(\vec{v})$ is also Gaussian, the integrals over velocity
in both cases can be performed analytically:
We use a mixture of three isotropic, zero-mean Gaussian distributions
\begin{equation}
  p(\vec{v}) = \sum_{m=1}^3 \, w_m \mathcal{N}(0, \sigma_{v,m}^2)
\end{equation}
with velocity dispersions $(\sigma_{v,1}, \sigma_{v,2}, \sigma_{v,3}) = (15, 30, 50)
~\kms$ and weights $(w_1,w_2,w_3) = (0.3, 0.55, 0.15)$
meant to encompass young thin disk stars to halo stars.
We derive the relevant expressions in Appendix~\ref{sec:appendix}.
After marginalizing over velocity, the likelihood integrands only depend on
distance; we numerically compute the integrals over the true distances to each
star in a pair, $d_1,d_2$, using Monte Carlo integration with $K$ samples from
the distance posterior pdfs assuming a uniform density prior
\citep{Bailer-Jones:2015aa}
\begin{equation}
  \int \, \dd d \, \tilde{L}(d) \, p(d\given\varpi) \approx
    \frac{1}{K} \, \sum_k^K \, \tilde{L}(d_k)
\end{equation}
where $\tilde{L}(d)$ is the velocity-marginalized likelihood function.
Through experimentation, we have found that $K=128$ samples are sufficient for
estimating the above integrals for stars with a wide range in parallax
signal-to-noise.
%TODO: SMOH: still unsure about this

@@

\begin{figure*}[htbp]
  \begin{center}
    \includegraphics[width=\textwidth]{figures/likelihoodratios.pdf}
  \end{center}
  \caption{%
    Justification of the likelihood ratio cut.
    On the top row, we show how the distribution of pairs on \absdvtan\ vs. separation changes
    in slices of $\ln \mathcal{L}_1 /\mathcal{L}_2$ (sequentially color-coded in blue).
    On the bottom row, we show how the distribution of $\ln \mathcal{L}_1 /\mathcal{L}_2$
    changes in slices of \absdvtan\ (sequentially color-coded in red).
    \label{fig:likelihoodratios}}
    \vspace{1em}
\end{figure*}

@@

Once we have identified candidate co-moving star pairs from the initial pair
sample, these pairs form an undirected graph where stars are nodes, and edges
between the nodes exist for co-moving pairs of stars.
A star may have multiple co-moving neighbors, and two stars may either
be directly or indirectly connected by a set of edges.
We divide the network into connected components\footnote{A connected component of an undirected graph $G$
is a subgraph of $G$ in which any two nodes are connected to each other by a path.},
and show the distribution of their sizes in Figure~\ref{fig:hist_ccsize}.
The most common are size 2 networks, which mean mutually exclusive co-moving
pairs. However, it is clear that there are many aggregates of co-moving stars
discovered by looking for co-moving pairs.
These aggregates are likely moving groups, OB associations, or star clusters.
There are 4555 connected components among 10,606 unique stars.
The maximum size of connected components is 151.
We show this largest network along with four randomly selected
examples of networks with sizes greater than 4 in Figure~\ref{fig:graphviz_examples}.
The largest network corresponds to the Pleiades open cluster.
One particular star in this network, TYC 1799-272-1, has 94 directly connected
co-moving neighbors, and is located at the center of the network.

@@

\begin{figure*}[htbp]
  \begin{center}
    \includegraphics[width=\textwidth]{figures/gjg_mepairs.pdf}
  \end{center}
  \caption{
    Color-magnitude diagrams of stars in networks of size 2 (i.e., mutually exclusive)
    in two separation bins.
    We connect each pair by a line on the left for pairs with separations smaller than
    1~pc.
    We show the distribution of all stars common in \tgas\ and \tmass\
    with parallax $S/N>16$ in blue for reference same as in Figure~\ref{fig:cmd_large}.
    \label{fig:cmd_me}}
\end{figure*}
